% -*- root: warwickthesis.tex -*-

Writing the acknowledgements is a truly challenging task as I want to express my gratitude to everyone who has helped shape my work and life over the years. Hence before I start, my sincerest apologies to anyone I did not mention by name.

Most importantly I would like to thank God for increasing my faculties and giving me the strength to finalise this work. Without His help I would not be where I am now as a person and in my research.

I also owe a debt of gratitude to \emph{Huzoor}, Hadhrat Mirza Masroor Ahmad (atba), for his invaluable counsel, he has always been a source of strength and has guided me through my life.

My parents, Mohammad Safdar Rana and Tahira Parveen Rana, have been a great inspiration and driving force for me.  I will be eternally grateful for the sacrifices that especially my mother made, to ensure we (her children)  have all the opportunities in life she never had the chance to pursue. I am immensely thankful for their prayers.  It is specifically for my mother's benefit that I have included a translation of these Acknowledgements in Urdu.

Each student and supervisor represents a unique pairing that you enter into almost blindly. I am one of the lucky ones, who found a supervisor that was always helpful, encouraging and above all motivational. I cannot thank Sach Mukherjee, my main supervisor, enough for the inspiration and occasional reminders when needed. Mario Nicodemi my secondary supervisor also provided invaluable suggestions and discussions.

My cohort at the Complexity Science DTC has always been helpful and supportive, unfortunately I didn't get the chance to spend much time at Warwick. I especially want to mention Leigh, Pantelis, Becky, Matt and Chris. I had the chance to spend two years of my PhD spend at the Netherlands Cancer Institute (NKI) on the eighth floor  with an amazing view of Amsterdam. I can't go without mentioning the support from the Mukherjee lab (Nic, Steven, Frank and Chris), their friendship and our unique lunch discussions. During my time there due to interactions with many Biologists I was able to further my understanding of biology itself as well as gain an appreciation for the different ways we approach questions and deal with problems and I thank everyone who helped me get there.

I am also obliged for the prayers and support of my friends, though I will not be able to name everyone here. I would like to mention Naveed who I shared a house with in Coventry for a year and who has been a great friend ever since. I thank Tauseef sahib for the encouragement and support over the years.  It would be remiss of me not to mention the members of the AMRA for the engaging discussions, which at times verged into the silly but were still great fun, among those were Muddassar, Azhaar, Shahzaib, Qasid, and Talha.

Deserving of special mention is my family  who I am extremely thankful to, for their encouragements and moral support. My two sisters Sadia and Farida, my brother in-law Mubashir as well as my younger brother Arslan. Farida also helped me to translate this into Urdu.

Lastly I want to mention someone very close to my hear who came into my life and changing me for the better, my wife, my dear friend Qurratul Ain. You have made me experience life in way I didn't know was possible. My thank also goes to her parents, Dr Samiullah Tahir and Faozia Tahir, and her two sisters for their support and prayers.
